\documentclass[11pt,a4paper]{article}
\usepackage[margin=0.85in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{caption}
\usepackage[hidelinks]{hyperref}
\usepackage[most,breakable]{tcolorbox}
\tcbuselibrary{breakable,skins}
\usepackage{fontspec}
\setmainfont{DejaVu Serif}
\setsansfont{DejaVu Sans}
\setmonofont{DejaVu Sans Mono}
\captionsetup{font=small,labelfont=bf,skip=4pt}
\setlist{leftmargin=1.5em,topsep=2pt}

%% Breakable problem card — prevents content cutoff across pages
\newtcolorbox{solcard}[3]{
  breakable, enhanced,
  colback=white, colframe=gray!50, arc=2pt,
  title={\textbf{#1}\hfill\textsf{\small #3\,pts}},
  fonttitle=\normalsize\sffamily\bfseries,
  before upper={\small\textit{#2}\par\smallskip\hrule height 0.4pt\smallskip},
  top=4pt, bottom=6pt, left=7pt, right=7pt,
  parbox=false,
}

\title{\textbf{Double dumbbell --- Solution}}
\author{\normalsize X21 (2021) — English translation}
\date{}
\begin{document}\maketitle\vspace{-0.5em}
\begin{solcard}{A1}{Find the angular velocity vector of the second rod at the initial moment of time. Express your answer in terms of $\vec\omega$.}{0.30}

At the moment the rotation begins, the ball at the end of the second rod is motionless. Therefore, the rods begin to rotate in opposite directions. Since the balls are at rest and the lengths of the rods are the same $$\vec{\omega}_2=-\vec{\omega}_1 $$

\end{solcard}

\begin{solcard}{A2}{At the initial moment of time, find the angular velocity vector $\vec\omega_{отн_0}$ of the second rod in the reference system associated with the first rod. Express your answer using $\vec\omega$}{0.20}

From the law of addition of angular velocities $$\vec{\omega}_{отн}=\vec{\omega}_2-\vec{\omega}_1 $$ Комбинируя equality с resultом partа $A1$ $$\vec{\omega}_{отн}=-2\vec{\omega} $$

\end{solcard}

\begin{solcard}{A3}{Find the potential energy of the load in the field of centrifugal force at a distance $r$ from the axis of rotation accurate to an arbitrary constant. Express your answer using $m$, $\omega$, $r$.}{0.50}

The potential energy can be obtained as the potential energy of a spring with a spring constant of $-m{\omega}^2$. Hence $$W_p=-\frac{m{\omega}^2r^2}{2}+C $$ In what follows, we will assume that the constant is equal to zero.

\end{solcard}

\begin{solcard}{A4}{Let the second rod rotate at an angle $\varphi$ relative to the first. Find its angular velocity $\omega_{отн}$ in this position. Express your answer using $\omega$ and $\varphi$.}{0.50}

As mentioned earlier, the Coriolis force does not change the kinetic energy of the system. Then the law of conservation of mechanical energy is satisfied in the system $$W_k+W_p=E $$ В initial moment времени $$2m{\omega}^2L^2=W_k $$ $$-2m{\omega}^2L^2=W_p $$ Thus $$E=0 $$ В произвольный moment времени $$W_k=\frac{m{\omega_{отн}}^2L^2}{2} $$ $$W_p=-2m{\omega}^2L^2\cos^2\left(\frac{\varphi}{2}\right) $$ Hence $$\vec{\omega}_{отн}=-2\vec{\omega}\cos\left(\frac{\varphi}{2}\right) $$

\end{solcard}

\begin{solcard}{A5}{Find the speed of the load in a fixed frame of reference. Express your answer in terms of $\omega$, $L$, $\varphi$.}{0.50}

One velocity component will be directed along the second rod, and the other will be perpendicular to it. We will receive $$v_1={\omega}L\sin(\varphi) $$ $$v_2={\omega}L(1+\cos(\varphi))-{\omega_{отн}}L $$ For квадрата скорости we get $$v^2={v_1}^2+{v_2}^2=(2\omega{L})^2\left(\left(\cos^2\left(\frac{\varphi}{2}\right)-\cos\left(\frac{\varphi}{2}\right)\right)^2+\sin^2\left(\frac{\varphi}{2}\right)\cos^2\left(\frac{\varphi}{2}\right)\right) $$ Hence после простых преобсований $$v=2{\omega}L\sqrt{2\left(\cos^2\left(\frac{\varphi}{2}\right)-\cos^3\left(\frac{\varphi}{2}\right)\right)} $$

\end{solcard}

\begin{solcard}{B1}{Let us denote the tension force of the rods as $T$ $$ $$ Find ускорения грузов $a_{1l}$,$a_{1p}$ в проекции на ось $OX$ and acceleration ball/sphereнира $a_2$ в проекции на ось $OY$.Answerы выtimesandте via/through $m$.$g$.$T$ and $\varphi$.}{0.40}

$$a_{1лx}=\frac{T\cos(\varphi)}{m} $$ $$a_{1пx}=-\frac{T\cos(\varphi)}{m} $$ $$a_{2y}=-\left(\frac{T\sin(\varphi)}{m}+g\right) $$

\end{solcard}

\begin{solcard}{B2}{In the indicated position, find the speed of the weights $v_1$ and the speed of the hinge $v_2$.}{0.60}

From the law of conservation of mechanical energy $$\frac{2m}{2}({v_1}^2+{v_2}^2)=2m\left(\frac{{v_0}^2}{2}-gL\sin(\varphi)\right) $$ Equation кинематической связи $${v_1}={v_2}\tan(\varphi) $$ Hence $$v_1=\sqrt{{v_0}^2-2gL\sin(\varphi)}\sin(\varphi) $$ $$v_2=\sqrt{{v_0}^2-2gL\sin(\varphi)}\cos(\varphi) $$

\end{solcard}

\begin{solcard}{B3}{Write down the equation of motion of the hinge relative to any of the masses, connecting $v_1$, $v_2$, $a_1$, $a_2$, $L$ and $\varphi$}{1.00}

The hinge moves in a circle relative to the loads. Hence $$\frac{v^2_{отн}}{L}={a_{1лx}}\cos(\varphi)-{a_{2y}}\sin(\varphi) $$ Since скорости ball/sphereнира and грузов перпендикулярны $$v^2_{отн}={v_1}^2+{v_2}^2 $$ and finally $$\frac{{v_1}^2+{v_2}^2}{L}={a_{1лx}}\cos(\varphi)-{a_{2y}}\sin(\varphi) $$

\end{solcard}

\begin{solcard}{B4}{Find the tension force $T$ from the equation obtained in step $B3$. Express the answer in terms of $m,{v_0},g,L,{\varphi}$}{0.70}

Combining the results of the first three points $$\left(\frac{{v_0}^2}{L}-2g\sin(\varphi)\right)=\frac{T}{m}+g\sin(\varphi) $$ hence $$T=m\left(\frac{{v_0}^2}{L}-3g\sin(\varphi)\right) $$

\end{solcard}

\begin{solcard}{B5}{Let us assume that at some moment the weights are lifted off the surface. $$ $$ Find value угла $\varphi_{neg}$ at this point in time.}{1.00}

Condition under which weights are lifted off the surface $$T\sin(\varphi)=mg $$ Substituting answer на part $B4$ $$m\left(\frac{{v_0}^2}{L}-3g\sin(\varphi)\right)\sin(\varphi)=mg $$ hence $$3\sin^2(\varphi)-\frac{{v_0}^2}{gL}\sin(\varphi)+1=0 $$ Let us find корни квадратного уравнения $$\sin(\varphi)=\frac{\frac{{v_0}^2}{gL}\pm\sqrt{\left(\frac{{v_0}^2}{gL}\right)^2-12}}{6} $$ Первым достигается меньший корень, а потому окончательный answer $$\sin(\varphi)=\frac{\frac{{v_0}^2}{gL}-\sqrt{\left(\frac{{v_0}^2}{gL}\right)^2-12}}{6} $$

\end{solcard}

\begin{solcard}{B6}{Find the minimum value $v_0$ at which the weights will come off the surface and the value of the angle $\varphi' $ at the moment of lifting}{0.50}

The discriminant of the quadratic equation must be non-negative, hence $$v_0\geq{\sqrt{2\sqrt{3}gL}} $$ Минимум достигается in the case of нулевого дискриминанта, therefore $$sin(\varphi_{max})=\frac{1}{\sqrt{3}} $$

\end{solcard}

\begin{solcard}{C1}{At what values ​​of $R$ does the collision of shells occur before their separation from the surface under the action of the rods? Express your answer using $L$,$\varphi_{отр}$,}{0.20}

Condition under which the collision occurs before the separation of $$R\geq{L}\cos(\varphi) $$

\end{solcard}

\begin{solcard}{C2}{Let immediately after the collision the projections of the speed of the load on the axes $OX$ and $OY$ be equal to $v_{1x}$ and $v_{1y}$, respectively. $$ $$ Express проекцию скорости ball/sphereнира на вертикальную ось $v_{2y}$ via/through $v_{1x}$,$v_{1y}$ and ${\varphi_0}$.}{1.00}

The load moves in a circle relative to the hinge, so $$\pm{v_{1x}}=(v_{1y}-v_{2y})\tan(\varphi_0) $$ $$v_{2y}=v_{1y}\pm{\frac{v_{1x}}{\tan(\varphi_0)}} $$

\end{solcard}

\begin{solcard}{C3}{Find the velocity of the hinge $v_2$ immediately after the impact. Express your answer using ${v_0},g,{L},R$}{1.30}

Let's solve the problem in the reference frame of the center of mass. Let $v_x$ and $v_y$ be the modulus of the relative speed of the loads and the modulus of the vertical velocity of the load relative to the hinge, respectively. Then the expression for kinetic energy in the reference frame of the center of mass is the following $$\frac{m{v_x}^2}{4}+\frac{m{v_y}^2}{2}=W_k $$ После соударения это also выполняется. $$ $$ From the equation кинематической связи $$\frac{v_x}{v_y}=tg(\varphi_0) $$ Ссу после соударения это соratio also выполняется. Therefore, from energy conservation it follows, что после соударения magnitudes относительных скоростей остаются постоянными. Это говорит о том, что грузы обмениваются друг с другом горизонтальными составляющими скоростей, а ball/sphereнир обменивается с ними вертикальными составляющими скоростей. Thus, for/at соударении ball/sphereнир останавливается. $$v_{2}=0 $$

\end{solcard}

\begin{solcard}{C4}{Find the modulus of the load velocities $v_1$ immediately after the collision. Express the answer in terms of ${v_0},g,{L},R$}{1.30}

All kinetic energy goes to the loads. From the law of conservation of energy we get the answer $$v_{1}=\sqrt{{v_0}^2-2g\sqrt{L^2-R^2}} $$

\end{solcard}

\end{document}