\documentclass[11pt,a4paper]{article}
% Shared preamble for RuPhO English problem statements (match T1 2026 style).
% Use: \input{latex/rupho_en_preamble.tex} after \documentclass.
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{caption}
\usepackage{tabularx}
\usepackage{hyperref}

\captionsetup{font=small,labelfont=bf,skip=6pt}

\setlist[itemize]{leftmargin=1.2cm}
\setlist[enumerate]{leftmargin=1.2cm}

% One outer box per subproblem: [id | statement | points] (no inner rules)
\newcommand{\problembox}[3]{%
  \par\addvspace{0.42em}%
  \noindent
  \setlength{\fboxsep}{7.5pt}%
  \setlength{\fboxrule}{0.95pt}%
  \fbox{%
    \begin{minipage}{\dimexpr\linewidth-2\fboxsep-2\fboxrule\relax}
    \begin{tabularx}{\linewidth}{@{}>{\raggedright\arraybackslash\bfseries}p{2.1em} @{\hspace{0.9em}} >{\raggedright\arraybackslash}X @{\hspace{0.9em}} >{\raggedleft\arraybackslash\bfseries}p{2.4em}@{}}
    #1 & #2 & #3
    \end{tabularx}
    \end{minipage}%
  }%
  \par\addvspace{0.32em}%
}

\title{\textbf{Bubble fluctuations}}
\author{\normalsize X18 (2018) --- English translation}
\date{}
\begin{document}
\maketitle

Consider an air bubble of radius $R$ immersed in water of density $\rho$. At equilibrium, the pressure inside and outside the bubble is $p_0$, if the surface tension of water is not taken into account. Consider air to be an ideal diatomic gas. After a slight disturbance, the bubble begins to oscillate radially with a certain frequency. The following steps will help you find this frequency. This can be done by representing the potential energy of an oscillating bubble as $Ax^2$, and the kinetic energy as $B\left(dx/dt\right)^2$, where $x$ is a small change in radius. Then the frequency will be $\omega=\sqrt{A/B}$.

\problembox{A1}{Find the restoring pressure force acting on a small area $dS$ of the bubble surface when its radius changes adiabatically from $R$ to $R+x$, ($x\ll{R}$).}{0.70}

\problembox{A2}{Find the energy required to change $x$ from $0$ to some small value $x_0$.}{1.00}

\problembox{A3}{The radial speed of water at the surface of the bubble is $dx/dt$. Given that water is incompressible, find the radial speed of water at a distance $r$ ($r>R$) from the center of the bubble.}{1.00}

\problembox{A4}{Find the total kinetic energy of water.}{1.00}

\problembox{A5}{If we neglect the air mass, the kinetic energy found in the previous paragraph will be the kinetic energy of the oscillating bubble. Find the vibration frequency.}{0.50}

Now let's take into account the influence of surface tension. The surface tension coefficient $\sigma$ is defined as the energy required to create a surface of unit area ($W=\sigma S$).

\problembox{B1}{Find the contribution of surface tension to the energy required to create a bubble of radius $R$ in water.}{0.20}

\problembox{B2}{Find the frequency of bubble oscillations, taking into account the influence of surface tension of water. Consider the pressure outside the bubble equal to $p_0$.}{1.60}

\vspace{1em}
\noindent\textit{Source:} \url{https://pho.rs/p/311}

\end{document}